---
title: "Comparing metrics for continuous attributes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{metrics_comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(divMetrics)
library(tidyverse)
library(GGally)
library(viridis)  # For color scales
library(ggrepel) # For text annotation in chart
library(ggtext)

```


## Introduction

This analysis aims to calculate and compare five diversity scores for teams of size 5 and 10, drawn from an age distribution ranging from 25 to 65. We will use the following diversity measures:

#. Blau's Index with 5-year binning
#. Blau's Index with 10-year binning
#. Coefficient of Variation (CV)
#. Standard Deviation (SD)
#. Coverage & Evenness Index (CEI)
#. Rao-Simpson Index with 5-year binning (RSI)
#. Rao-Simpson Index with 10-year binning (RSI)

## Generate Team Data

To generate teams with a full spectrum of diversity, we introduce a diversity parameter sampled from a Beta distribution for each team. This parameter controls the probability of team members being sampled from a narrow versus a uniform distribution, thus resulting in a low/high diversity team.

```{r generate-teams, message=FALSE}
set.seed(123)  # Set seed for reproducibility
ages <- seq(25, 65, by = 1)

# Function to generate team ages with controlled diversity
generate_team <- function(size, n_teams, alpha = 2, beta = 5) {
  p_diversity <- rbeta(n_teams, alpha, beta)  # Diversity parameter for each team
  
  teams_list <- vector("list", n_teams)
  
  for (i in 1:n_teams) {
    p <- p_diversity[i]
    
    # Define the probability of selecting from a narrow range vs. the entire age range
    prob_narrow <- p
    prob_uniform <- 1 - p
    
    # Define a narrow age range (e.g., mean age ± 5 years)
    mean_age <- sample(30:60, 1)
    narrow_range <- (mean_age - 5):(mean_age + 5)
    narrow_range <- narrow_range[narrow_range >= 25 & narrow_range <= 65]
    
    # Define probabilities
    probs <- rep(prob_uniform / length(ages), length(ages))
    probs[narrow_range - 24] <- prob_narrow / length(narrow_range)
    
    # Sample team members
    team_ages <- sample(ages, size = size, replace = TRUE, prob = probs)
    
    teams_list[[i]] <- tibble(
      team_id = paste0("Team_", size, "_", i),
      age = team_ages,
      team_size = size
    )
  }
  
  bind_rows(teams_list)
}

# Create teams of size 5 and 10 with enhanced diversity simulation
teams_size_5 <- generate_team(5, 5000) %>% mutate(team_size = 5)
teams_size_10 <- generate_team(10, 5000) %>% mutate(team_size = 10)

# Combine all teams
teams <- bind_rows(teams_size_5, teams_size_10)
```


## Calculate Diversity Scores

### Calculate Scores

For clarity, we calculate each metric separately and then join them. Alternatively, you could use `compute_all_metrics()` for a more concise approach.

```{r calculate-scores}
# Calculate each metric separately for clarity
div_scores <- teams %>%
  group_by(team_id, team_size) %>%
  summarise(
    CEI = compute_CEI(age, range = c(min(teams$age), max(teams$age))),
    Rao_Simpson_5 = compute_Rao(age, bins = seq(25, 65, by = 5)),
    Rao_Simpson_10 = compute_Rao(age, bins = seq(25, 65, by = 10)),
    Blau_5 = compute_Blau(age, bins = seq(25, 65, by = 5)),
    Blau_10 = compute_Blau(age, bins = seq(25, 65, by = 10)),
    CV = compute_cv(age),
    SD = compute_sd(age),
    .groups = "drop"
  )
```

## Visualize Diversity Score Distributions

To compare the distribution of all diversity measures across different team sizes, we will use violin plots faceted by team size. This approach, using *normalized* scores, provides a clear visualization of the distribution shape and spread for each measure.

```{r visualize-violins, fig.width=12, fig.height=8}
# Reshape data to long format for easier plotting
div_long <- div_scores %>%
  mutate(across(CEI:SD, ~c(scale(.x, center = FALSE)))) %>% 
  pivot_longer(cols = CEI:SD, names_to = "Measure", values_to = "Value") %>% 
  mutate(team_size = paste("Teams of", team_size) %>% factor(levels = c("Teams of 5", "Teams of 10")))

# Create violin plots
ggplot(div_long, aes(x = Measure, y = Value, fill = Measure)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  facet_wrap(~ team_size, scales = "free") +
  scale_fill_viridis(discrete = TRUE, option = "C") +
  #theme_minimal() +
  labs(
    title = "Distribution of Diversity Measures by Team Size ...",
    subtitle = "... divided by their standard deviation for comparabiity",
    x = "Diversity Measure",
    y = "Score",
    fill = "Measure"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Correlation Analysis

```{r correlation-analysis, fig.width=8, fig.height=6}
div_scores %>%
  select(CEI:SD) %>%
  timesaveR::cor_matrix() %>% 
  timesaveR::report_cor_table()
```

Key points to note:
- CEI is uniformly related to both variety and variance-based measures, with correlation coefficients close to .7
- Blau's index strongly depends on the width of bins, with the two specifications only correlating at .75. This effect is much weaker for the Rao-Simpson index, which correlates at .93.

## Illustrative comparison

```{r illustrative-comparison}
example_teams <- bind_rows(
  tibble(ages = c(20, 30, 40, 50, 60), team = "A"),
  tibble(ages = c(20, 60, 60, 60, 60), team = "B"),
  tibble(ages = c(20, 20, 20, 20, 60), team = "C"),
  tibble(ages = c(29, 30, 49, 50, 60), team = "D"),
  tibble(ages = c(30, 31, 50, 51, 61), team = "E"),
  tibble(ages = c(20, 20, 20, 29, 30), team = "F")
)

# Calculate metrics for each team
example_scores <- example_teams %>%
  group_by(team) %>%
  summarise(
    CEI = compute_CEI(ages, range = c(min(example_teams$ages), max(example_teams$ages))),
    Rao_5 = compute_Rao(ages, bins = seq(20, 65, by = 5)),
    Rao_10 = compute_Rao(ages, bins = seq(20, 70, by = 10)),
    Blau_5 = compute_Blau(ages, bins = seq(20, 65, by = 5)),
    Blau_10 = compute_Blau(ages, bins = seq(20, 70, by = 10)),
    CV = compute_cv(ages),
    SD = compute_sd(ages),
    team_members = paste(ages, collapse = ", "),
    .groups = "drop"
  )

# Quick comparison option using built-in plotting function:
# plot_metric_comparison(example_scores,
#                       metrics = c("CV", "SD", "Blau_5", "Blau_10", "Rao_5", "Rao_10", "CEI"),
#                       group_var = "team",
#                       group_members_var = "team_members",
#                       title = "Comparison of Normalised Diversity Indices")

# For this analysis, we create a custom visualization with detailed styling:

example_scores_long <- example_scores %>%
  pivot_longer(
    cols = c(
      CV, SD, Blau_5, Blau_10,
      Rao_5, Rao_10, CEI
    ),
    names_to = "Index",
    values_to = "Score"
  ) %>%
  group_by(Index) %>%
  mutate(Score = scale(Score, center = FALSE)) %>%
  ungroup()

# Reorder factor levels for Index
example_scores_long$Index <- factor(example_scores_long$Index,
  levels = c(
    "CV", "SD", "Blau_5", "Blau_10"
    , "Rao_5", "Rao_10", "CEI"
  )
)

# Define line types and colors to highlight clusters:
line_types <- c(
  "A" = "solid",
  "B" = "dashed", "C" = "dashed",
  "D" = "dotdash", "E" = "dotdash",
  "F" = "longdash"
)

team_colors <- c(
  "A" = "black",
  "B" = "darkred",  "C" = "red",
  "D" = "darkblue", "E" = "blue",
  "F" = "forestgreen"
)

p <- ggplot(
  example_scores_long,
  aes(
    x = Index, y = Score, group = team,
    linetype = team, color = team
  )
) +
  geom_line(aes(size = team)) +
  geom_point() +
  scale_linetype_manual(values = line_types) +
  scale_color_manual(values = team_colors) +
  scale_size_manual(values = c(
    "A" = 1.2, "B" = 0.8, "C" = 0.8,
    "D" = 0.8, "E" = 0.8, "F" = 0.8
  )) +
  labs(
    title = "Comparison of Normalised Diversity Indices",
    x = "", y = ""
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.margin = margin(t = 20, r = 10, b = 20, l = 50)
  ) +
  scale_x_discrete(
    expand = expansion(mult = c(0.2, 0.05)),
    labels = c(
      "CV" = "CV",
      "SD" = "SD",
      "Blau_5" = "Blau (5)",
      "Blau_10" = "Blau (10)"
    )
  )

label_data <- example_scores_long %>%
  filter(Index == "CV") %>%
  select(Index, team, Score, team_members)

p + geom_text_repel(
  data = label_data,
  aes(label = paste0(team, ": ", team_members), x = Index, y = Score),
  nudge_x = -.75,
  vjust = 0.5,
  size = 3
) +
  scale_y_continuous(position = "right")

```

Key points to note:

- Teams B and C diverge hugely on the CV, but are identical on all other indices. This highlights that the CV's conceptualisation of diversisty as *disparity* is distinct and requires justification.
- Teams D and E substantially differ on the indices that depend on binning - even though E is simply one year older than D. This discontinuity highlights the issue with binning.
- Within this example, each operationalisation of diversity results in a differnt ranking of teams. This highlights that the choice of measure matters.


## Divergence between the measures

Next we explore *where* diversity measures diverge. For that, we plot the associations and identify the greatest residuals. These cases can then help identify which measure best identifies diverse teams.

### Comparing Blau's Index and CEI


```{r divergence_blau5, fig.width=10, fig.height=6}
p <- div_scores %>% 
  filter(team_size == 5) %>%
  ggplot(aes(x = CEI, y = Blau_5)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(alpha = 0.5) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  theme_minimal() +
  labs(
    title = "Blau's Index (5-year Binning) vs. CEI",
    subtitle = "Teams of 5",
    x = "Coverage & Evenness Index (CEI)",
    y = "Blau's Index (5-year Binning)",
  ) 

p %>% 
  ggExtra::ggMarginal(p, type="histogram")


outliers <- div_scores %>% 
  filter(team_size == 5) %>%
  group_by(Blau_5 = round(Blau_5, 2)) %>% 
  mutate(count = n()) %>% 
  arrange(CEI) %>%
  # select the largest and smallest CEI per group
  filter(CEI == max(CEI, na.rm = TRUE) | CEI == min(CEI, na.rm = TRUE)) %>% 
  mutate(score_type = ifelse(CEI == max(CEI, na.rm = TRUE), "max", "min")) %>% 
  select(CEI, Blau_5, group_members, team_size, score_type, count) %>% 
  ungroup() %>% 
  group_by(Blau_5) %>%
  arrange(CEI) %>% 
  filter(row_number() == 1 | row_number() == n(), n() > 1) %>% # exclude if there is only 1 team (v rare), otherwise filter 1 min and 1 max
  ungroup()
```


```{r fig.height=4, fig.width=10}
# Plot CEI range for most common Blau value in each .1 range

  outliers_parsed <- outliers %>%
    mutate(blau_group = floor(Blau_5 * 10)) %>% 
    group_by(blau_group) %>%
    filter(count == max(count), Blau_5 > .1) %>% 
    mutate(members = map(group_members, ~ timesaveR::line_to_vector(.x, return = "vector") %>% as.numeric())) %>%
    unnest(cols = members)

  # 2) Create separate summary frames for min and max teams
  annotation_min <- outliers_parsed %>%
    filter(score_type == "min") %>%
    group_by(Blau_5, score_type) %>%
    summarise(
      CEI   = first(CEI),
      CEI_C = compute_CEI(members, range = range, return = "C"),
      CEI_E = compute_CEI(members, range = range, return = "E"),
      x_center = mean(members),  # center of the distribution
      .groups = "drop"
    )

  annotation_max <- outliers_parsed %>%
    filter(score_type == "max") %>%
    group_by(Blau_5, score_type) %>%
    summarise(
      CEI   = first(CEI),
      CEI_C = compute_CEI(members, range = range, return = "C"),
      CEI_E = compute_CEI(members, range = range, return = "E"),
      x_center = mean(members),
      .groups = "drop"
    )

  my_colors <- c(
  "min" = "#918821",   # e.g., a teal/blue
  "max" = "#8297ed"    # e.g., a reddish color
)
  
  ggplot() +
    geom_hline(yintercept = 0, color = "black") +
    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "min"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "down"
    ) +

    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "max"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "up" 
    ) +

    geom_text(
      data = annotation_min,
      aes(
        x = mean(range),
        y = -0.4,  
        label = paste0(
          "CEI = ", round(CEI, 2),
          " (C = ", round(CEI_C, 2),
          ", E = ", round(CEI_E, 2), ")"
        ),
        color = score_type
      ),
      size = 3.5,
      inherit.aes = FALSE
    ) +

    geom_text(
      data = annotation_max,
      aes(
        x = mean(range),
        y = 0.4,  
        label = paste0(
          "CEI = ", round(CEI, 2),
          " (C = ", round(CEI_C, 2),
          ", E = ", round(CEI_E, 2), ")"
        ),
                color = score_type
      ),
      size = 3.5,
      inherit.aes = FALSE
    ) +

    facet_wrap(~ Blau_5, ncol = 3) +
    scale_x_continuous(limits = range, breaks = seq(range[1], range[2], by = 5)) +
    # Avoid clipping if text goes beyond panel
    coord_cartesian(clip = "off") +
      theme_void() +
  theme(
    strip.background = element_rect(fill = "grey85", color = NA),
    strip.text = element_text(color = "black"),
    legend.position = "none",
        plot.subtitle = element_markdown()

  ) +
    labs(
      title = "Deviations between CEI and Blau in 5-member teams",
    subtitle = paste0(
      "Showing the <b><span style='color:", my_colors["max"], ";'>most</span></b> ",
      "and <b><span style='color:", my_colors["min"], ";'>least</span> </b>",
      "CEI-diverse team for common Blau values<br>&nbsp;"
    )
    ) +
      scale_fill_manual(values = my_colors) +
  scale_color_manual(values = my_colors) 


```

### Comparing the Coefficient of Variation and CEI

```{r divergence_blau5, fig.width=10, fig.height=6}
p <- div_scores %>% 
  filter(team_size == 5) %>%
  ggplot(aes(x = CEI, y = CV)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(alpha = 0.5) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  theme_minimal() +
  labs(
    title = "CV vs. CEI",
    subtitle = "Teams of 5",
    x = "Coverage & Evenness Index (CEI)",
    y = "Coefficient of Variation",
  ) 

p %>% 
  ggExtra::ggMarginal(p, type="histogram")


outliers <- div_scores %>%
  filter(team_size == 5) %>%
  group_by(CV = round(2*CV, 1)/2, team_size) %>% 
  mutate(count = n()) %>% 
  arrange(CV) %>%
  # select the largest and smallest CEI per group
  filter(CEI == max(CEI, na.rm = TRUE) | CEI == min(CEI, na.rm = TRUE)) %>% 
  mutate(score_type = ifelse(CEI == max(CEI, na.rm = TRUE), "max", "min")) %>% 
  select(CEI, CV, group_members, team_size, score_type, count) %>% 
  ungroup() %>% 
  group_by(CV) %>%
  arrange(CEI) %>%
  filter(row_number() == 1 | row_number() == n(), n() > 1) %>% # exclude if there is only 1 team (v rare), otherwise filter 1 min and 1 max
  ungroup()
```


```{r fig.height=4, fig.width=10}
# Plot CEI range for CV values rounded to the nearest .05

  outliers_parsed <- outliers %>%
    mutate(members = map(group_members, ~ timesaveR::line_to_vector(.x, return = "vector") %>% as.numeric())) %>%
    unnest(cols = members)

  # 2) Create separate summary frames for min and max teams
  annotation_min <- outliers_parsed %>%
    filter(score_type == "min") %>%
    group_by(CV, score_type) %>%
    summarise(
      CEI   = first(CEI),
      CEI_C = compute_CEI(members, range = range, return = "C"),
      CEI_E = compute_CEI(members, range = range, return = "E"),
      x_center = mean(members),  # center of the distribution
      .groups = "drop"
    )

  annotation_max <- outliers_parsed %>%
    filter(score_type == "max") %>%
    group_by(CV, score_type) %>%
    summarise(
      CEI   = first(CEI),
      CEI_C = compute_CEI(members, range = range, return = "C"),
      CEI_E = compute_CEI(members, range = range, return = "E"),
      x_center = mean(members),
      .groups = "drop"
    )

  my_colors <- c(
  "min" = "#918821",   # e.g., a teal/blue
  "max" = "#8297ed"    # e.g., a reddish color
)
  
  ggplot() +
    geom_hline(yintercept = 0, color = "black") +
    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "min"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "down"
    ) +

    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "max"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "up" 
    ) +

    geom_text(
      data = annotation_min,
      aes(
        x = mean(range),
        y = -0.4,  
        label = paste0(
          "CEI = ", round(CEI, 2),
          " (C = ", round(CEI_C, 2),
          ", E = ", round(CEI_E, 2), ")"
        ),
        color = score_type
      ),
      size = 3.5,
      inherit.aes = FALSE
    ) +

    geom_text(
      data = annotation_max,
      aes(
        x = mean(range),
        y = 0.4,  
        label = paste0(
          "CEI = ", round(CEI, 2),
          " (C = ", round(CEI_C, 2),
          ", E = ", round(CEI_E, 2), ")"
        ),
                color = score_type
      ),
      size = 3.5,
      inherit.aes = FALSE
    ) +

    facet_wrap(~ CV, ncol = 3) +
    scale_x_continuous(limits = range, breaks = seq(range[1], range[2], by = 5)) +
    # Avoid clipping if text goes beyond panel
    coord_cartesian(clip = "off") +
      theme_void() +
  theme(
    strip.background = element_rect(fill = "grey85", color = NA),
    strip.text = element_text(color = "black"),
    legend.position = "none",
        plot.subtitle = element_markdown()

  ) +
    labs(
      title = "Deviations between CEI and CV in 5-member teams",
    subtitle = paste0(
      "Showing the <b><span style='color:", my_colors["max"], ";'>most</span></b> ",
      "and <b><span style='color:", my_colors["min"], ";'>least</span> </b>",
      "CEI-diverse team for common CV values<br>&nbsp;"
    )
    ) +
      scale_fill_manual(values = my_colors) +
  scale_color_manual(values = my_colors) 


```


## Summary

In this analysis, we generated teams of two different sizes (5 and 10) with a comprehensive range of diversity levels by controlling the diversity parameter through a Beta distribution. Five diversity metrics—Even Spread Index (ESI), Blau's Index (with 5-year and 10-year binning), Coefficient of Variation (CV), and Standard Deviation (SD)—were computed for each team.

**Key Findings:**

- **Distribution of Diversity Measures:** Violin plots revealed the distribution and spread of each diversity measure across different team sizes. While Blau's index is very clustered around a low number of unique values, all other distributions are smooth and roughly normal.

- **Correlation Analysis:** The correlation matrix indicated strong correlations between certain measures (e.g., CV and SD) while others showed weaker relationships. This suggests that while some diversity metrics capture similar aspects of diversity, others may provide unique insights.

- **Measures Divergence:** A comparison of 6 specific teams highlighted that each index can result in a unique ordering, so that the choice matters. Similarly, looking at the residuals between CEI and Blau's Index or CV, we identified the teams where the measures diverged the most - and would argue that in all these cases, CEI indicates an appropriate difference between the teams.
