---
title: "Comparing metrics for continuous attributes"
output: 
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
vignette: >
  %\VignetteIndexEntry{Comparing metrics for continuous attributes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(divMetrics)
library(dplyr)
library(ggplot2)
library(purrr)
library(GGally)
library(viridis)  # For color scales
library(ggrepel) # For text annotation in chart
library(ggtext)
library(tidyr)

# Load internal helper functions 
# (replacements for timesaveR functions, see https://lukaswallrich.github.io/timesaveR/)
cor_matrix <- divMetrics:::cor_matrix
report_cor_table <- divMetrics:::report_cor_table
line_to_vector <- divMetrics:::line_to_vector
```


## Introduction

This analysis aims to calculate and compare five diversity scores for teams of size 5 and 10, drawn from an age distribution ranging from 20 to 70. We will use the following diversity measures:

#. Blau's Index with 5-year binning
#. Blau's Index with 10-year binning
#. Coefficient of Variation (CV)
#. Standard Deviation (SD)
#. Coverage & Evenness Index (CEI)
#. Gini Mean Difference (GMD; continuous Rao equivalent)

## Generate Team Data

To generate teams with a full spectrum of diversity, we use a hybrid approach where a continuous diversity parameter directly controls the width and spread of team compositions. This creates a smooth continuum from very homogeneous teams through moderate diversity to highly diverse teams including bimodal and skewed configurations. This aims to capture the range of team configurations observed empirically.

```{r generate-teams, message=FALSE}
set.seed(123)  # Set seed for reproducibility

# Generate teams with varying diversity levels using a simpler structural approach
# The diversity parameter (p) influences team structure and spread, but is not directly coupled to metrics
# - For unimodal teams: normal distribution around a random center; width scales with p
# - For bimodal teams: two normal clusters with separation/width scaled by p; split can vary (~50/50 ± 15%)
# - For core_outlier: tight core + 1 far outlier; offset scales with p
# This avoids overfitting to metric logic and produces more natural variation
generate_team_simple <- function(size, n_teams, bounds = c(20, 70), alpha = 1, beta = 1) {
  span <- diff(bounds)
  clamp <- function(x) pmin(pmax(x, bounds[1]), bounds[2])
  rnorm_clamped <- function(n, mean, sd) round(clamp(rnorm(n, mean, sd)))

  p_div <- rbeta(n_teams, alpha, beta)  # diversity level in [0,1]
  teams_list <- vector("list", n_teams)

  for (i in seq_len(n_teams)) {
    p <- p_div[i]
    center <- runif(1, bounds[1], bounds[2])

    # Simple shape choice: low p → unimodal; higher p → mix
    if (p < 0.30) {
      type <- "unimodal"
    } else {
      type <- sample(c("unimodal", "bimodal", "core_outlier"),
                     size = 1, prob = c(0.4, 0.3, 0.3))
    }

    if (type == "unimodal") {
      # Width scales with p; normal draw adds natural clumpiness
      sd <- 0.8 + 10 * p
      ages <- rnorm_clamped(size, center, sd)

    } else if (type == "bimodal") {
      # Separation grows with p; clusters narrower when far apart
      sep <- 6 + 24 * p
      sd <- 1 + 4 * (1 - p)
      w <- 0.5 + runif(1, -0.15, 0.15)      # ~50/50 but not exact
      n1 <- max(1, round(size * w)); n2 <- max(1, size - n1)
      if (n1 + n2 > size) { if (runif(1) < 0.5) n1 <- n1 - 1 else n2 <- n2 - 1 }
      c1 <- clamp(center - sep/2); c2 <- clamp(center + sep/2)
      ages <- c(rnorm_clamped(n1, c1, sd), rnorm_clamped(n2, c2, sd))

    } else { # core_outlier: tight core + 1 far member
      n_core <- size - 1
      sd_core <- 0.8 + 2 * p
      core <- rnorm_clamped(n_core, center, sd_core)
      offset <- sample(c(-1, 1), 1) * (12 + 18 * p)
      outlier <- rnorm_clamped(1, center + offset, sd = 1.5)
      ages <- c(core, outlier)
    }

    teams_list[[i]] <- tibble(team_id = paste0("Team_", size, "_", i),
                              age = ages,
                              team_size = size,
                              diversity_profile = type,
                              diversity_prop = p)
  }
  dplyr::bind_rows(teams_list)
}

# Create teams of size 5 and 10 with controlled diversity
teams_size_5 <- generate_team_simple(5, 2500)
teams_size_10 <- generate_team_simple(10, 2500)

# Combine all teams
teams <- bind_rows(teams_size_5, teams_size_10)
```


## Calculate Diversity Scores

### Calculate Scores

For clarity, we calculate each metric separately and then join them. Alternatively, you could use `compute_all_metrics()` for a more concise approach.

```{r calculate-scores}
# Calculate each metric separately for clarity
div_scores <- teams %>%
  group_by(team_id, team_size) %>%
  summarise(
    CEI = compute_CEI(age, range = c(20, 70)),
    Blau_5 = compute_Blau(age, bins = seq(20, 70, by = 5)),
    Blau_10 = compute_Blau(age, bins = seq(20, 70, by = 10)),
    GMD = compute_GMD(age),
    CV = compute_cv(age),
    SD = compute_sd(age),
    group_members = paste(age, collapse = ","),
    .groups = "drop"
  )
```

## Visualize Diversity Score Distributions

To compare the distribution of all diversity measures across different team sizes, we use violin plots faceted by team size. This approach, using *normalized* scores, provides a clear visualization of the distribution shape and spread for each measure. This highlights that Blau index values are not-continuous and heavily influenced by binning choices, while CEI, CV, and SD show smooth distributions.

```{r visualize-violins, fig.width=12, fig.height=8}
# Reshape data to long format for easier plotting
div_long <- div_scores %>%
  mutate(across(CEI:SD, ~c(scale(.x, center = FALSE)))) %>% 
  pivot_longer(cols = CEI:SD, names_to = "Measure", values_to = "Value") %>% 
  mutate(team_size = paste("Teams of", team_size) %>% factor(levels = c("Teams of 5", "Teams of 10")))

# Create violin plots
ggplot(div_long, aes(x = Measure, y = Value, fill = Measure)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  facet_wrap(~ team_size, scales = "free") +
  scale_fill_viridis(discrete = TRUE, option = "C") +
  #theme_minimal() +
  labs(
    title = "Distribution of Diversity Measures by Team Size ...",
    subtitle = "... divided by their standard deviation for comparabiity",
    x = "Diversity Measure",
    y = "Score",
    fill = "Measure"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Correlation Analysis

```{r correlation-analysis, fig.width=8, fig.height=6}
div_scores %>%
  select(CEI:SD) %>%
  cor_matrix() %>%
  report_cor_table()
```

Key points to note:

- CEI is related to both variety and variance-based measures, with correlation coefficients typically around .7–.85.
- Blau's index depends on the width and alignment of bins; continuous GMD avoids these binning artefacts.

However, the specific correlation values depend on the simulation parameters. As will be shown below, different team configurations can lead to substantial divergence between measures - and the extent to which these teams appear in real-world samples will influence divergences between measures.

## Illustrative comparison

```{r illustrative-comparison, fig.width=10, fig.height=6}
example_teams <- bind_rows(
  tibble(ages = c(20, 30, 40, 50, 60), team = "A"),
  tibble(ages = c(20, 60, 60, 60, 60), team = "B"),
  tibble(ages = c(20, 20, 20, 20, 60), team = "C"),
  tibble(ages = c(29, 30, 49, 50, 60), team = "D"),
  tibble(ages = c(30, 31, 50, 51, 61), team = "E"),
  tibble(ages = c(20, 20, 20, 29, 30), team = "F")
)

# Calculate metrics for each team
example_scores <- example_teams %>%
  group_by(team) %>%
  summarise(
    CEI = compute_CEI(ages, range = c(20, 70)),
    Blau_5 = compute_Blau(ages, bins = seq(20, 70, by = 5)),
    Blau_10 = compute_Blau(ages, bins = seq(20, 70, by = 10)),
    GMD = compute_GMD(ages),
    CV = compute_cv(ages),
    SD = compute_sd(ages),
    team_members = paste(ages, collapse = ", "),
    .groups = "drop"
  )

# Quick comparison option using built-in plotting function:
# plot_metric_comparison(example_scores,
#                       metrics = c("CV", "SD", "Blau_5", "Blau_10", "GMD", "CEI"),
#                       group_var = "team",
#                       group_members_var = "team_members",
#                       title = "Comparison of Normalised Diversity Indices")

# For this analysis, we create a custom visualization with detailed styling:

example_scores_long <- example_scores %>%
  pivot_longer(
    cols = c(
      CV, SD, Blau_5, Blau_10,
      GMD, CEI
    ),
    names_to = "Index",
    values_to = "Score"
  ) %>%
  group_by(Index) %>%
  mutate(Score = scale(Score, center = FALSE)) %>%
  ungroup() %>%
  mutate(
    indicator_type = factor(
      if_else(Index %in% c("CV", "SD", "Blau_5", "Blau_10"),
              "Established Indicators",
              "Hybrid Indicators"),
      levels = c("Established Indicators", "Hybrid Indicators")
    )
  )

# Reorder factor levels for Index and create numeric x positions
# This allows lines to connect across the two indicator groups
example_scores_long <- example_scores_long %>%
  mutate(
    Index = factor(Index,
      levels = c(
        "CV", "SD", "Blau_5", "Blau_10",
        "GMD", "CEI"
      )
    ),
    # Map indices to numeric positions with a gap between groups
    x_pos = case_when(
      Index == "CV" ~ 1,
      Index == "SD" ~ 2,
      Index == "Blau_5" ~ 3,
      Index == "Blau_10" ~ 4,
      Index == "GMD" ~ 6,
      Index == "CEI" ~ 8
    )
  )

# Define line types and colors to highlight clusters:
line_types <- c(
  "A" = "solid",
  "B" = "dashed", "C" = "dashed",
  "D" = "dotdash", "E" = "dotdash",
  "F" = "longdash"
)

team_colors <- c(
  "A" = "black",
  "B" = "darkred",  "C" = "red",
  "D" = "darkblue", "E" = "blue",
  "F" = "forestgreen"
)

# Create the base plot with numeric x positions to allow line connections
p <- ggplot(
  example_scores_long,
  aes(
    x = x_pos, y = Score, group = team,
    linetype = team, color = team
  )
) +
  geom_line(aes(linewidth = team)) +
  geom_point() +
  # Add subtle background sections to distinguish indicator types
  annotate("rect", xmin = 0.5, xmax = 4.5, ymin = -Inf, ymax = Inf,
           alpha = 0.1, fill = "grey") +
  annotate("rect", xmin = 5.5, xmax = 8.5, ymin = -Inf, ymax = Inf,
           alpha = 0.1, fill = "white") +
  # Add vertical separator line
  annotate("vline", xintercept = 4.7, linetype = "dashed",
           color = "grey70", size = 0.3) +
  scale_linetype_manual(values = line_types) +
  scale_color_manual(values = team_colors) +
  scale_linewidth_manual(values = c(
    "A" = 1.2, "B" = 0.8, "C" = 0.8,
    "D" = 0.8, "E" = 0.8, "F" = 0.8
  )) +
  labs(
    title = "Comparison of Normalised Diversity Indices",
    x = "", y = ""
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.margin = margin(t = 20, r = 10, b = 20, l = 50),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  scale_x_continuous(
    breaks = c(1, 2, 3, 4, 6, 8),
    labels = c("CV", "SD", "Blau (5)", "Blau (10)", "GMD", "CEI"),
    expand = expansion(mult = c(0.08, 0.08))
  )

# Add section labels
p <- p +
  annotate("text", x = 2.5, y = Inf, label = "Established Indicators",
           vjust = 1.5, size = 3, color = "grey50") +
  annotate("text", x = 6.5, y = Inf, label = "Hybrid Indicators",
           vjust = 1.5, size = 3, color = "grey50")

# Add labels on the left (at CV)
label_data_left <- example_scores_long %>%
  filter(Index == "CV") %>%
  select(x_pos, team, Score, team_members)

p <- p + geom_text_repel(
  data = label_data_left,
  aes(label = paste0(team, ": ", team_members), x = x_pos, y = Score),
  nudge_x = -0.6,
  vjust = 0.5,
  size = 3,
  direction = "y",
  segment.size = 0.3,
  segment.alpha = 0.5
)

# Add labels on the right (at CEI)
label_data_right <- example_scores_long %>%
  filter(Index == "CEI") %>%
  select(x_pos, team, Score, team_members)

p <- p + geom_text_repel(
  data = label_data_right,
  aes(label = paste0(team, ": ", team_members), x = x_pos, y = Score),
  nudge_x = 0.6,
  vjust = 0.5,
  size = 3,
  direction = "y",
  segment.size = 0.3,
  segment.alpha = 0.5
)

p

```

Key points to note:

- Teams B and C diverge hugely on the CV, but are identical on all other indices (including GMD). This highlights that CV’s scaling by the mean captures a different aspect than pure distance-based spread.
- Teams D and E substantially differ on the indices that depend on binning - even though E is simply one year older than D. This discontinuity highlights the issue with binning.
- Within this example, each operationalisation of diversity results in a different ranking of teams. This highlights that the choice of measure matters (though in practice, many teams will be similar across measures).


## Divergence between the measures

Next we explore *where* diversity measures diverge. For that, we plot the associations and identify the greatest residuals. These cases can then help identify which measure best identifies diverse teams.

### Comparing Blau's Index and CEI

Blau's index and CEI are correlated, yet each level of Blau's index corresponds to a range of CEI values. We explore these deviations below.

```{r divergence_blau5, fig.width=10, fig.height=6}
p <- div_scores %>% 
  filter(team_size == 5) %>%
  ggplot(aes(x = CEI, y = Blau_5)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(alpha = 0.5) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  theme_minimal() +
  labs(
    title = "Blau's Index (5-year Binning) vs. CEI",
    subtitle = "Teams of 5",
    x = "Coverage & Evenness Index (CEI)",
    y = "Blau's Index (5-year Binning)",
  ) 

p %>% 
  ggExtra::ggMarginal(p, type="histogram")


outliers <- div_scores %>% 
  filter(team_size == 5) %>%
  group_by(Blau_5 = round(Blau_5, 2)) %>% 
  mutate(count = n()) %>% 
  arrange(CEI) %>%
  # select the largest and smallest CEI per group
  filter(CEI == max(CEI, na.rm = TRUE) | CEI == min(CEI, na.rm = TRUE)) %>% 
  mutate(score_type = ifelse(CEI == max(CEI, na.rm = TRUE), "max", "min")) %>% 
  select(CEI, Blau_5, group_members, team_size, score_type, count) %>% 
  ungroup() %>% 
  group_by(Blau_5) %>%
  arrange(CEI) %>% 
  filter(row_number() == 1 | row_number() == n(), n() > 1) %>% # exclude if there is only 1 team (v rare), otherwise filter 1 min and 1 max
  ungroup()
```

The chart below shows, for each common Blau value the team with the highest and lowest CEI. This highlights how much CEI can vary even when Blau is constant.

```{r fig.height=6, fig.width=10}
# Plot CEI range for most common Blau value in each .1 range

  outliers_parsed <- outliers %>%
    mutate(blau_group = floor(Blau_5 * 10)) %>%
    group_by(blau_group) %>%
    filter(count == max(count), Blau_5 > .1) %>%
    mutate(members = map(group_members,
                         ~ line_to_vector(.x, return = "vector"))) %>%
    unnest(cols = members)

  # 2) Create separate summary frames for min and max teams
  annotation_min <- outliers_parsed %>%
    filter(score_type == "min") %>%
    group_by(Blau_5, score_type) %>%
    summarise(
      CEI   = first(CEI),
      CEI_C = compute_CEI(members, range = c(20, 70), return = "C"),
      CEI_E = compute_CEI(members, range = c(20, 70), return = "E"),
      x_center = mean(members),  # center of the distribution
      .groups = "drop"
    )

  annotation_max <- outliers_parsed %>%
    filter(score_type == "max") %>%
    group_by(Blau_5, score_type) %>%
    summarise(
      CEI   = first(CEI),
      CEI_C = compute_CEI(members, range = c(20, 70), return = "C"),
      CEI_E = compute_CEI(members, range = c(20, 70), return = "E"),
      x_center = mean(members),
      .groups = "drop"
    )

  my_colors <- c(
  "min" = "#918821",   # e.g., a teal/blue
  "max" = "#8297ed"    # e.g., a reddish color
)
  
  ggplot() +
    geom_hline(yintercept = 0, color = "black") +
    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "min"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "down"
    ) +

    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "max"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "up" 
    ) +

    geom_text(
      data = annotation_min,
      aes(
        x = mean(c(20, 70)),
        y = -0.4,  
        label = paste0(
          "CEI = ", round(CEI, 2),
          " (C = ", round(CEI_C, 2),
          ", E = ", round(CEI_E, 2), ")"
        ),
        color = score_type
      ),
      size = 3.5,
      inherit.aes = FALSE
    ) +

    geom_text(
      data = annotation_max,
      aes(
        x = mean(c(20, 70)),
        y = 0.4,  
        label = paste0(
          "CEI = ", round(CEI, 2),
          " (C = ", round(CEI_C, 2),
          ", E = ", round(CEI_E, 2), ")"
        ),
                color = score_type
      ),
      size = 3.5,
      inherit.aes = FALSE
    ) +

    facet_wrap(~ Blau_5, ncol = 3) +
    scale_x_continuous(limits = c(20, 70), breaks = seq(20, 70, by = 5)) +
    # Avoid clipping if text goes beyond panel
    coord_cartesian(clip = "off") +
      theme_void() +
  theme(
    strip.background = element_rect(fill = "grey85", color = NA),
    strip.text = element_text(color = "black"),
    legend.position = "none",
        plot.subtitle = element_markdown()

  ) +
    labs(
      title = "Deviations between CEI and Blau in 5-member teams",
    subtitle = paste0(
      "Showing the <b><span style='color:", my_colors["max"], ";'>most</span></b> ",
      "and <b><span style='color:", my_colors["min"], ";'>least</span> </b>",
      "CEI-diverse team for common Blau values<br>&nbsp;"
    )
    ) +
      scale_fill_manual(values = my_colors) +
  scale_color_manual(values = my_colors) 


```

### Comparing the Coefficient of Variation and CEI

```{r divergence_cv, fig.width=10, fig.height=6}
p <- div_scores %>% 
  filter(team_size == 5) %>%
  ggplot(aes(x = CEI, y = CV)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(alpha = 0.5) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  theme_minimal() +
  labs(
    title = "CV vs. CEI",
    subtitle = "Teams of 5",
    x = "Coverage & Evenness Index (CEI)",
    y = "Coefficient of Variation",
  ) 

p %>% 
  ggExtra::ggMarginal(p, type="histogram")


outliers <- div_scores %>%
  filter(team_size == 5) %>%
  group_by(CV = round(2*CV, 1)/2, team_size) %>% 
  mutate(count = n()) %>% 
  arrange(CV) %>%
  # select the largest and smallest CEI per group
  filter(CEI == max(CEI, na.rm = TRUE) | CEI == min(CEI, na.rm = TRUE)) %>% 
  mutate(score_type = ifelse(CEI == max(CEI, na.rm = TRUE), "max", "min")) %>% 
  select(CEI, CV, group_members, team_size, score_type, count) %>% 
  ungroup() %>% 
  group_by(CV) %>%
  arrange(CEI) %>%
  filter(row_number() == 1 | row_number() == n(), n() > 1) %>% # exclude if there is only 1 team (v rare), otherwise filter 1 min and 1 max
  ungroup()
```

The chart below shows, for each rounded CV value the team with the highest and lowest CEI. This highlights that particularly for intermediate CV values, CEI can vary widely (and towards high CV values, the relationship becomes reversed as evenness drops).

```{r fig.height=4, fig.width=10}
# Plot CEI range for CV values rounded to the nearest .05

  outliers_parsed <- outliers %>%
    mutate(members = map(group_members,
                         ~ line_to_vector(.x, return = "vector"))) %>%
    unnest(cols = members)

  # 2) Create separate summary frames for min and max teams
  annotation_min <- outliers_parsed %>%
    filter(score_type == "min") %>%
    group_by(CV, score_type) %>%
    summarise(
      CEI   = first(CEI),
      CEI_C = compute_CEI(members, range = c(20, 70), return = "C"),
      CEI_E = compute_CEI(members, range = c(20, 70), return = "E"),
      x_center = mean(members),  # center of the distribution
      .groups = "drop"
    )

  annotation_max <- outliers_parsed %>%
    filter(score_type == "max") %>%
    group_by(CV, score_type) %>%
    summarise(
      CEI   = first(CEI),
      CEI_C = compute_CEI(members, range = c(20, 70), return = "C"),
      CEI_E = compute_CEI(members, range = c(20, 70), return = "E"),
      x_center = mean(members),
      .groups = "drop"
    )

  my_colors <- c(
  "min" = "#918821",   # e.g., a teal/blue
  "max" = "#8297ed"    # e.g., a reddish color
)
  
  ggplot() +
    geom_hline(yintercept = 0, color = "black") +
    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "min"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "down"
    ) +

    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "max"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "up" 
    ) +

    geom_text(
      data = annotation_min,
      aes(
        x = mean(c(20, 70)),
        y = -0.4,  
        label = paste0(
          "CEI = ", round(CEI, 2),
          " (C = ", round(CEI_C, 2),
          ", E = ", round(CEI_E, 2), ")"
        ),
        color = score_type
      ),
      size = 3.5,
      inherit.aes = FALSE
    ) +

    geom_text(
      data = annotation_max,
      aes(
        x = mean(c(20, 70)),
        y = 0.4,  
        label = paste0(
          "CEI = ", round(CEI, 2),
          " (C = ", round(CEI_C, 2),
          ", E = ", round(CEI_E, 2), ")"
        ),
                color = score_type
      ),
      size = 3.5,
      inherit.aes = FALSE
    ) +

    facet_wrap(~ CV, ncol = 3) +
    scale_x_continuous(limits = c(20, 70), breaks = seq(20, 70, by = 5)) +
    # Avoid clipping if text goes beyond panel
    coord_cartesian(clip = "off") +
      theme_void() +
  theme(
    strip.background = element_rect(fill = "grey85", color = NA),
    strip.text = element_text(color = "black"),
    legend.position = "none",
        plot.subtitle = element_markdown()

  ) +
    labs(
      title = "Deviations between CEI and CV in 5-member teams",
    subtitle = paste0(
      "Showing the <b><span style='color:", my_colors["max"], ";'>most</span></b> ",
      "and <b><span style='color:", my_colors["min"], ";'>least</span> </b>",
      "CEI-diverse team for common CV values<br>&nbsp;"
    )
    ) +
      scale_fill_manual(values = my_colors) +
  scale_color_manual(values = my_colors) 


```


## Summary

In this analysis, we generated teams of two different sizes (5 and 10) with a comprehensive range of member distributions. Five diversity metrics—Even Spread Index (ESI), Blau's Index (with 5-year and 10-year binning), Coefficient of Variation (CV), and Standard Deviation (SD)—were computed for each team.

**Key Findings:**

- **Distribution of Diversity Measures:** Violin plots revealed the distribution and spread of each diversity measure across different team sizes. While Blau's index is very clustered around a low number of unique values, all other distributions are smooth and roughly normal.

- **Correlation Analysis:** The correlation matrix indicated strong correlations between certain measures (e.g., CV and SD) while others showed weaker relationships. This suggests that while some diversity metrics capture similar aspects of diversity, others may provide unique insights.

- **Measures Divergence:** A comparison of 6 specific teams highlighted that each index can result in a unique ordering, so that the choice matters. Similarly, looking at the residuals between CEI and Blau's Index or CV, we identified the teams where the measures diverged the most - and would argue that in all these cases, CEI indicates an appropriate difference between the teams.
