---
title: "Comparing metrics for continuous attributes"
output:
  html_document:
    toc: true
    toc_depth: 2
    code_folding: hide
  pdf_document:
    latex_engine: xelatex
    keep_tex: false
classoption: oneside
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhead[C]{\textit{Code available in Rmd source and HTML version}}
---

```{r, include = FALSE}
is_pdf <- knitr::opts_knit$get("rmarkdown.pandoc.to") == "latex"

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = !is_pdf  # Hide code in PDF, show in HTML
)
```

```{r setup}
library(divMetrics)
library(dplyr)
library(ggplot2)
library(purrr)
library(GGally)
library(viridis)  # For color scales
library(ggrepel) # For text annotation in chart
library(ggtext)
library(tidyr)
library(gt)  # For tables

# Load internal helper functions
cor_matrix <- divMetrics:::cor_matrix
report_cor_table <- divMetrics:::report_cor_table
line_to_vector <- divMetrics:::line_to_vector
```


## Introduction

This analysis aims to calculate and compare diversity scores for teams of
size 5 and 10, drawn from an age distribution ranging from 20 to 70. We will
use the following diversity measures:

#. Blau's Index with 5-year binning
#. Blau's Index with 10-year binning
#. Coefficient of Variation (CV)
#. Standard Deviation (SD)
#. Coverage & Evenness Index (CEI)

## Generate Team Data

```{r generate-teams-text, echo=FALSE, results='asis'}
if (is_pdf) {
  cat("Team diversity is generated using a hybrid approach where a continuous",
      "diversity parameter directly controls the width and spread of team",
      "compositions. This creates a smooth continuum from very homogeneous",
      "teams (narrow age ranges of 2-3 years) through moderate diversity",
      "(wider unimodal distributions) to highly diverse teams including",
      "bimodal (two separated clusters) and skewed (unequal cluster sizes)",
      "configurations. This approach ensures that the underlying diversity",
      "parameter has a strong, direct relationship with measured diversity",
      "while still capturing the range of team configurations observed",
      "empirically.\n")
}
```

```{r generate-teams, message=FALSE}
set.seed(123)  # Set seed for reproducibility

# Generate teams with varying diversity levels
# width_prop controls span; teams can be unimodal or two-cluster
generate_team <- function(size, n_teams, bounds = c(20, 70),
                          p_two_cluster = 0.3,
                          width_shape1 = 0.8, width_shape2 = 0.8) {
  min_age   <- bounds[1]
  max_age   <- bounds[2]
  total_span <- max_age - min_age

  clamp_round <- function(x) {
    round(pmin(pmax(x, min_age), max_age))
  }

  teams <- vector("list", n_teams)

  for (i in seq_len(n_teams)) {

    # 1. Draw a width (span) on [0, total_span]
    # mostly narrow, some wide
    width_prop <- rbeta(1, width_shape1, width_shape2)
    width <- max(1, width_prop * total_span)  # at least 1 year

    # 2. Place that interval somewhere within [min_age, max_age]
    start <- runif(1, min_age, max_age - width)
    low   <- start
    high  <- start + width

    # 3. Decide on unimodal vs two-cluster
    two_cluster <- runif(1) < p_two_cluster

    if (!two_cluster) {
      ages <- runif(size, low, high)
      shape <- "unimodal"

    } else {
      # Two clusters with flexible relative sizes and separation
      # Cluster sizes vary; extreme splits produce "outlier-like"
      # teams naturally
      # encourages unequal splits
      w  <- rbeta(1, 0.8, 0.8)
      n1 <- max(1, round(size * w))
      n2 <- max(1, size - n1)
      if (n1 + n2 > size) {
        if (runif(1) < 0.5) n1 <- n1 - 1 else n2 <- n2 - 1
      }

      # Position clusters inside the span
      # c1 somewhere in the left part, c2 somewhere in the right part
      c1 <- runif(1, low + 0.1*(high - low), low + 0.4*(high - low))
      c2 <- runif(1, low + 0.6*(high - low), low + 0.9*(high - low))

      # cluster SD scales with width (tighter for small spans)
      sd_cluster <- 0.1 * (high - low)

      ages <- c(
        rnorm(n1, c1, sd_cluster),
        rnorm(n2, c2, sd_cluster)
      )

      shape <- "two_cluster"
    }
    ages <- clamp_round(ages)

    teams[[i]] <- tibble(
      team_id           = paste0("Team_", size, "_", i),
      age               = ages,
      team_size         = size,
      diversity_profile = shape,
      diversity_prop    = width_prop
    )
  }

  dplyr::bind_rows(teams)
}

# Create teams of size 5 and 10 with controlled diversity
teams_size_5 <- generate_team(5, 2500)
teams_size_10 <- generate_team(10, 2500)

# Combine all teams
teams <- bind_rows(teams_size_5, teams_size_10)
```


## Calculate Diversity Scores

```{r calculate-scores-text, echo=FALSE, results='asis'}
if (is_pdf) {
  cat("For each team, we calculate the five diversity metrics listed above.")
} else {
  cat("### Calculate Scores\n\n")
  cat("For clarity, we calculate each metric separately and then join them.",
      "Alternatively, you could use `compute_all_metrics()` for a more",
      "concise approach.\n")
}
```

```{r calculate-scores}
# Calculate each metric separately for clarity
div_scores <- teams %>%
  group_by(team_id, team_size) %>%
  summarise(
    CEI = compute_CEI(age, range = c(20, 70)),
    Blau_5 = compute_Blau(age, bins = seq(20, 70, by = 5)),
    Blau_10 = compute_Blau(age, bins = seq(20, 70, by = 10)),
    CV = compute_cv(age),
    SD = compute_sd(age),
    group_members = paste(age, collapse = ","),
    .groups = "drop"
  )
```

## Visualize Diversity Score Distributions

To compare the distribution of all diversity measures across different team
sizes, we use violin plots faceted by team size. This approach, using
*normalized* scores, provides a clear visualization of the distribution
shape and spread for each measure. This highlights that Blau index values
are not-continuous and heavily influenced by binning choices, while CEI, CV,
and SD show smooth distributions.

```{r visualize-violins, fig.width=12, fig.height=8}
# Reshape data to long format for easier plotting
div_long <- div_scores %>%
  # Normalize by dividing by SD
  mutate(across(CEI:SD, ~c(scale(.x, center = FALSE)))) %>%
  pivot_longer(
    cols = CEI:SD,
    names_to = "Measure",
    values_to = "Value"
  ) %>%
  mutate(
    team_size = paste("Teams of", team_size) %>%
      factor(levels = c("Teams of 5", "Teams of 10"))
  )

# Create violin plots
ggplot(div_long, aes(x = Measure, y = Value, fill = Measure)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  facet_wrap(~ team_size, scales = "free") +
  scale_fill_viridis(discrete = TRUE, option = "C") +
  labs(
    title = "Distribution of Diversity Measures by Team Size",
    subtitle = "Divided by their standard deviation for comparability",
    x = "Diversity Measure",
    y = "Score",
    fill = "Measure"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Correlation Analysis

```{r correlation-analysis, fig.width=8, fig.height=6}
div_scores %>%
  select(CEI:SD) %>%
  cor_matrix() %>%
  report_cor_table()
```

Key points to note:

- CEI is related to both variety and variance-based measures, with
  correlation coefficients typically around .7–.85.
- Blau's index depends on the width and alignment of bins; continuous CEI
  avoids these binning artefacts.

However, the specific correlation values depend on the simulation
parameters. As will be shown below, different team configurations can lead
to substantial divergence between measures - and the extent to which these
teams appear in real-world samples will influence divergences between
measures.

## Illustrative comparison

```{r illustrative-comparison, fig.width=10, fig.height=6}
example_teams <- bind_rows(
  tibble(ages = c(20, 30, 40, 50, 60), team = "A"),
  tibble(ages = c(20, 60, 60, 60, 60), team = "B"),
  tibble(ages = c(20, 20, 20, 20, 60), team = "C"),
  tibble(ages = c(24, 29, 30, 49, 50), team = "D"),
  tibble(ages = c(25, 30, 31, 50, 51), team = "E"),
  tibble(ages = c(30, 30, 40, 50, 50), team = "F"),
  tibble(ages = c(20, 20, 40, 60, 60), team = "G")
)

# Calculate metrics for each team
example_scores <- example_teams %>%
  group_by(team) %>%
  summarise(
    CEI = compute_CEI(ages, range = c(20, 70)),
    Blau_5 = compute_Blau(ages, bins = seq(20, 70, by = 5)),
    Blau_10 = compute_Blau(ages, bins = seq(20, 70, by = 10)),
    CV = compute_cv(ages),
    SD = compute_sd(ages),
    team_members = paste(ages, collapse = ", "),
    .groups = "drop"
  )

# Quick comparison option using built-in plotting function:
# plot_metric_comparison(example_scores,
#   metrics = c("CV", "SD", "Blau_5", "Blau_10", "CEI"),
#   group_var = "team",
#   group_members_var = "team_members",
#   title = "Comparison of Normalised Diversity Indices")

# For this analysis, we create a custom visualization with detailed
# styling:

example_scores_long <- example_scores %>%
  pivot_longer(
    cols = c(CV, SD, Blau_5, Blau_10, CEI),
    names_to = "Index",
    values_to = "Score"
  ) %>%
  group_by(Index) %>%
  mutate(Score = scale(Score, center = FALSE)) %>%
  ungroup() %>%
  mutate(
    indicator_type = factor(
      if_else(
        Index %in% c("CV", "SD", "Blau_5", "Blau_10"),
        "Established Indicators",
        "Coverage & Evenness Index"
      ),
      levels = c("Established Indicators", "Coverage & Evenness Index")
    )
  )

# Reorder factor levels for Index and create numeric x positions
# This allows lines to connect across the two indicator groups
example_scores_long <- example_scores_long %>%
  mutate(
    Index = factor(
      Index,
      levels = c("CV", "SD", "Blau_5", "Blau_10", "CEI")
    ),
    # Map indices to numeric positions with a gap between groups
    x_pos = case_when(
      Index == "CV" ~ 1,
      Index == "SD" ~ 2,
      Index == "Blau_5" ~ 3,
      Index == "Blau_10" ~ 4,
      Index == "CEI" ~ 6
    )
  )

# Define line types and colors to highlight clusters:
line_types <- c(
  "A" = "solid",
  "B" = "dashed", "C" = "dashed",
  "D" = "dotdash", "E" = "dotdash",
  "F" = "longdash", "G" = "longdash"
)

team_colors <- c(
  "A" = "black",
  "B" = "darkred",  "C" = "red",
  "D" = "darkblue", "E" = "blue",
  "F" = "forestgreen", "G" = "#004000"
)

# Create the base plot with numeric x positions to allow line connections
p <- ggplot(
  example_scores_long,
  aes(
    x = x_pos, y = Score, group = team,
    linetype = team, color = team
  )
) +
  geom_line(aes(linewidth = team)) +
  geom_point() +
  # Add subtle background sections to distinguish indicator types
  annotate("rect", xmin = 0.5, xmax = 4.5, ymin = -Inf, ymax = Inf,
           alpha = 0.1, fill = "white") +
  annotate("rect", xmin = 5.5, xmax = 7.2, ymin = -Inf, ymax = Inf,
           alpha = 0.1, fill = "#807f7f") +
  scale_linetype_manual(values = line_types) +
  scale_color_manual(values = team_colors) +
  scale_linewidth_manual(values = c(
    "A" = 0.6, "B" = 0.6, "C" = 0.6,
    "D" = 0.6, "E" = 0.6, "F" = 0.6, "G" = 0.6
  )) +
  labs(
    title = "Comparison of Normalised Diversity Indices",
    subtitle = paste(
      "Lines show normalized scores for seven teams",
      "across different metrics"
    ),
    x = "", y = ""
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.margin = margin(t = 20, r = 10, b = 20, l = 50),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  scale_x_continuous(
    breaks = c(1, 2, 3, 4, 6),
    labels = c("CV", "SD", "Blau (5)", "Blau (10)", "CEI"),
    expand = expansion(mult = c(0.08, 0.08))
  )

# Add labels on the left (at CV)
label_data_left <- example_scores_long %>%
  filter(Index == "CV") %>%
  select(x_pos, team, Score, team_members)

p <- p + geom_text_repel(
  data = label_data_left,
  aes(label = paste0(team, ": ", team_members), x = x_pos, y = Score),
  nudge_x = -0.6,
  vjust = 0.5,
  size = 3,
  direction = "y",
  segment.size = 0.3,
  segment.alpha = 0.5
)

# Add labels on the right (at CEI)
label_data_right <- example_scores_long %>%
  filter(Index == "CEI") %>%
  select(x_pos, team, Score, team_members)

p <- p + geom_text_repel(
  data = label_data_right,
  aes(label = paste0(team, ": ", team_members), x = x_pos, y = Score),
  nudge_x = 0.6,
  vjust = 0.5,
  size = 3,
  direction = "y",
  segment.size = 0.3,
  segment.alpha = 0.5
)

p

```

Key points to note:

- Teams B and C diverge hugely on the CV, but are identical on all other
  indices. This highlights that CV's scaling by the mean captures a unique
  aspect that needs to be theoretically justified.
- Teams D and E substantially differ on the indices that depend on binning
  - even though E is simply one year older than D. This discontinuity
  highlights a key issue with binning.
- Within this example, each operationalisation of diversity results in a
  different ranking of teams. This highlights that the choice of measure
  matters (though in practice, many teams will be similar across measures).


## Divergence between the measures

Next we explore *where* diversity measures diverge. For that, we plot the
associations and identify the greatest residuals. These cases can then help
identify which measure best identifies diverse teams.

### Comparing Blau's Index and CEI

Blau's index and CEI are correlated, yet each level of Blau's index
corresponds to a range of CEI values. We explore these deviations below.

```{r divergence_blau5, fig.width=10, fig.height=6}
p <- div_scores %>%
  filter(team_size == 5) %>%
  ggplot(aes(x = CEI, y = Blau_5)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(alpha = 0.5) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  theme_minimal() +
  labs(
    title = "CEI vs Blau's Index (5-year Binning)",
    subtitle = "Teams of 5",
    x = "Coverage & Evenness Index (CEI)",
    y = "Blau's Index (5-year Binning)"
  )

p %>%
  ggExtra::ggMarginal(p, type = "histogram")


outliers <- div_scores %>% 
  filter(team_size == 5) %>%
  group_by(Blau_5 = round(Blau_5, 2)) %>% 
  mutate(count = n()) %>% 
  arrange(CEI) %>%
  # select the largest and smallest CEI per group
  filter(CEI == max(CEI, na.rm = TRUE) | CEI == min(CEI, na.rm = TRUE)) %>% 
  mutate(score_type = ifelse(CEI == max(CEI, na.rm = TRUE), "max", "min")) %>% 
  select(CEI, Blau_5, group_members, team_size, score_type, count) %>% 
  ungroup() %>% 
  group_by(Blau_5) %>%
  arrange(CEI) %>% 
  filter(row_number() == 1 | row_number() == n(), n() > 1) %>% # exclude if there is only 1 team (v rare), otherwise filter 1 min and 1 max
  ungroup()
```

```{r divergence_blau_examples, fig.height=5, fig.width=10, include=FALSE}
# Plot CEI range for most common Blau value in each .1 range

  outliers_parsed <- outliers %>%
    mutate(blau_group = floor(Blau_5 * 10)) %>%
    group_by(blau_group) %>%
    filter(count == max(count), Blau_5 > .1) %>%
    mutate(members = map(group_members,
                         ~ line_to_vector(.x, return = "vector"))) %>%
    unnest(cols = members)

  # Calculate fold-change for each Blau bin
  fold_changes <- outliers_parsed %>%
    group_by(Blau_5) %>%
    summarise(
      max_CEI = max(CEI),
      min_CEI = min(CEI),
      fold_change = max_CEI / pmax(min_CEI, 0.01),  # Avoid division by zero
      .groups = "drop"
    )

  # 2) Create separate summary frames for min and max teams
  annotation_min <- outliers_parsed %>%
    filter(score_type == "min") %>%
    group_by(Blau_5, score_type) %>%
    summarise(
      CEI   = first(CEI),
      x_center = mean(members),  # center of the distribution
      .groups = "drop"
    )

  annotation_max <- outliers_parsed %>%
    filter(score_type == "max") %>%
    group_by(Blau_5, score_type) %>%
    summarise(
      CEI   = first(CEI),
      x_center = mean(members),
      .groups = "drop"
    )

  my_colors <- c(
    "min" = "#918821",
    "max" = "#8297ed"
  )

 p_div_blau <- ggplot() +
    geom_hline(yintercept = 0, color = "black") +
    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "min"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "down",
      dotsize = 2
    ) +

    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "max"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "up",
      dotsize = 2
    ) +

    # CEI label for min (simplified)
    geom_text(
      data = annotation_min,
      aes(
        x = mean(c(20, 70)),
        y = -0.45,
        label = paste0("CEI = ", round(CEI, 2)),
        color = score_type
      ),
      size = 4.5,
      fontface = "bold",
      inherit.aes = FALSE
    ) +

    # CEI label for max (simplified)
    geom_text(
      data = annotation_max,
      aes(
        x = mean(c(20, 70)),
        y = 0.45,
        label = paste0("CEI = ", round(CEI, 2)),
        color = score_type
      ),
      size = 4.5,
      fontface = "bold",
      inherit.aes = FALSE
    ) +

    # Add fold-change annotation
    geom_text(
      data = fold_changes,
      aes(
        x = mean(c(20, 70)),
        y = -0.7,
        label = paste0(round(fold_change, 1), "× variation")
      ),
      size = 3.5,
      color = "black",
      fontface = "italic",
      inherit.aes = FALSE
    ) +

    facet_wrap(~ Blau_5, ncol = 2) +
    scale_x_continuous(limits = c(20, 70), breaks = seq(20, 70, by = 10)) +
    # Avoid clipping if text goes beyond panel
    coord_cartesian(clip = "off") +
    theme_void() +
    theme(
      strip.background = element_rect(fill = "grey85", color = NA),
      strip.text = element_text(color = "black", size = 11, face = "bold"),
      legend.position = "none",
      plot.subtitle = element_markdown(),
      plot.margin = margin(10, 10, 20, 10)
    ) +
    labs(
      title = "Deviations between CEI and Blau in 5-member teams",
      subtitle = paste0(
        "Showing the <b><span style='color:", my_colors["max"], ";'>most</span></b> ",
        "and <b><span style='color:", my_colors["min"], ";'>least</span></b> ",
        "CEI-diverse team for common Blau values"
      )
    ) +
    scale_fill_manual(values = my_colors) +
    scale_color_manual(values = my_colors)

p_div_blau
```

```{r blau_cei_breakdown, include=FALSE}
# Create a table showing CEI components for the 12 teams shown in the Blau chart

# Get unique team data (one row per team/score_type combination)
blau_table_data <- outliers %>%
  mutate(blau_group = floor(Blau_5 * 10)) %>%
  group_by(blau_group) %>%
  filter(count == max(count), Blau_5 > .1) %>%
  ungroup() %>%
  select(Blau_5, CEI, group_members, score_type) %>%
  distinct() %>%
  arrange(Blau_5, desc(score_type))

# Calculate C and E components for each team
blau_table_data <- blau_table_data %>%
  rowwise() %>%
  mutate(
    ages = list(line_to_vector(group_members, return = "vector") %>% as.numeric()),
    C = compute_CEI(ages, range = c(20, 70), return = "C"),
    E = compute_CEI(ages, range = c(20, 70), return = "E"),
    .keep = "unused"
  ) %>%
  ungroup() %>%
  mutate(
    Blau_Group = round(Blau_5, 2),
    Type = ifelse(score_type == "max", "High CEI", "Low CEI"),
    CEI = round(CEI, 3),
    C = round(C, 3),
    E = round(E, 3)
  ) %>%
  select(Blau_Group, Type, CEI, C, E) %>%
  group_by(Blau_Group) %>%
  arrange(desc(Type), .by_group = TRUE)

# Create gt table
blau_table_data %>%
  gt(groupname_col = "Blau_Group") %>%
  tab_header(
    title = "Coverage & Evenness Index (CEI) Components",
    subtitle = "Broken down by Coverage (C) and Evenness (E) for the 12 teams shown in Blau divergence chart"
  ) %>%
  cols_label(
    Type = "Type",
    CEI = "CEI",
    C = "Coverage (C)",
    E = "Evenness (E)"
  ) %>%
  fmt_number(
    columns = c(CEI, C, E),
    decimals = 3
  ) %>%
  tab_options(
    row_group.as_column = FALSE,
    container.overflow.x = TRUE
  )
```

### Comparing the Coefficient of Variation and CEI

```{r divergence_cv, fig.width=10, fig.height=6}
p <- div_scores %>%
  filter(team_size == 5) %>%
  ggplot(aes(x = CEI, y = CV)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(alpha = 0.5) +
  scale_color_viridis(discrete = TRUE, option = "D") +
  theme_minimal() +
  labs(
    title = "CEI vs Coefficient of Variation",
    subtitle = "Teams of 5",
    x = "Coverage & Evenness Index (CEI)",
    y = "Coefficient of Variation"
  )

p %>%
  ggExtra::ggMarginal(p, type = "histogram")


outliers <- div_scores %>%
  filter(team_size == 5) %>%
  group_by(CV = round(2*CV, 1)/2, team_size) %>% 
  mutate(count = n()) %>% 
  arrange(CV) %>%
  # select the largest and smallest CEI per group
  filter(CEI == max(CEI, na.rm = TRUE) | CEI == min(CEI, na.rm = TRUE)) %>% 
  mutate(score_type = ifelse(CEI == max(CEI, na.rm = TRUE), "max", "min")) %>% 
  select(CEI, CV, group_members, team_size, score_type, count) %>% 
  ungroup() %>% 
  group_by(CV) %>%
  arrange(CEI) %>%
  filter(row_number() == 1 | row_number() == n(), n() > 1) %>% # exclude if there is only 1 team (v rare), otherwise filter 1 min and 1 max
  ungroup()
```

```{r divergence_cv_examples, fig.height=5, fig.width=10, include=FALSE}
# Plot CEI range for selected CV values

  outliers_parsed <- outliers %>%
    # Keep only specific CV bins
    filter(CV %in% c(0.1, 0.2, 0.25, 0.3, 0.4, 0.5)) %>%
    mutate(members = map(group_members, ~ line_to_vector(.x, return = "vector") %>% as.numeric())) %>%
    unnest(cols = members) %>%
    # Add CV label (numeric value only)
    mutate(
      CV_label = CV
    )

  # Calculate fold-change for each CV bin
  fold_changes <- outliers_parsed %>%
    group_by(CV) %>%
    summarise(
      max_CEI = max(CEI),
      min_CEI = min(CEI),
      fold_change = max_CEI / pmax(min_CEI, 0.01),  # Avoid division by zero
      .groups = "drop"
    ) %>%
    left_join(outliers_parsed %>% select(CV, CV_label) %>% distinct(), by = "CV")

  # 2) Create separate summary frames for min and max teams
  annotation_min <- outliers_parsed %>%
    filter(score_type == "min") %>%
    group_by(CV) %>%
    summarise(
      CEI   = first(CEI),
      x_center = mean(members),
      score_type = first(score_type),
      .groups = "drop"
    ) %>%
    distinct(CV, .keep_all = TRUE) %>%
    # Add CV_label so faceting assigns each label to the right panel
    left_join(outliers_parsed %>% select(CV, CV_label) %>% distinct(), by = "CV")

  annotation_max <- outliers_parsed %>%
    filter(score_type == "max") %>%
    group_by(CV) %>%
    summarise(
      CEI   = first(CEI),
      x_center = mean(members),
      score_type = first(score_type),
      .groups = "drop"
    ) %>%
    distinct(CV, .keep_all = TRUE) %>%
    # Add CV_label so faceting assigns each label to the right panel
    left_join(outliers_parsed %>% select(CV, CV_label) %>% distinct(), by = "CV")

  my_colors <- c(
    "min" = "#918821",   
    "max" = "#8297ed"    
  )
  
  p_div_cv <- ggplot() +
    geom_hline(yintercept = 0, color = "black") +
    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "min"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "down",
      dotsize = 2
    ) +

    geom_dotplot(
      data = outliers_parsed %>% filter(score_type == "max"),
      aes(x = members, fill = score_type),
      binwidth = 1,
      stackdir = "up",
      dotsize = 2
    ) +

    # CEI label for min (removed C & E)
    geom_text(
      data = annotation_min,
      aes(
        x = mean(c(20, 70)),
        y = -0.45,  
        label = paste0("CEI = ", round(CEI, 2)),
        color = score_type
      ),
      size = 4.5,
      fontface = "bold",
      inherit.aes = FALSE
    ) +

    # CEI label for max (removed C & E)
    geom_text(
      data = annotation_max,
      aes(
        x = mean(c(20, 70)),
        y = 0.45,  
        label = paste0("CEI = ", round(CEI, 2)),
        color = score_type
      ),
      size = 4.5,
      fontface = "bold",
      inherit.aes = FALSE
    ) +

    # Add fold-change annotation
    geom_text(
      data = fold_changes,
      aes(
        x = mean(c(20, 70)),
        y = -0.7,
        label = paste0(round(fold_change, 1), "× variation")
      ),
      size = 3.5,
      color = "black",
      fontface = "italic",
      inherit.aes = FALSE
    ) +

    facet_wrap(~ CV_label, nrow = 3) +
    scale_x_continuous(limits = c(20, 70), breaks = seq(20, 70, by = 10)) +
    coord_cartesian(clip = "off") +
    theme_void() +
    theme(
      strip.background = element_rect(fill = "grey85", color = NA),
      strip.text = element_text(color = "black", size = 11, face = "bold"),
      legend.position = "none",
      plot.subtitle = element_markdown(),
      plot.margin = margin(10, 10, 20, 10)
    ) +
    labs(
      title = "Deviations between CEI and CV in 5-member teams",
      subtitle = paste0(
        "Showing the <b><span style='color:", my_colors["max"], ";'>most</span></b> ",
        "and <b><span style='color:", my_colors["min"], ";'>least</span></b> ",
        "CEI-diverse team for selected CV values (±0.025)"
      ),
      caption = "Each dot represents one team member's age (20-70)."
    ) +
    scale_fill_manual(values = my_colors) +
    scale_color_manual(values = my_colors)

    p_div_cv
```

```{r cv_cei_breakdown, include=FALSE}
# Create a table showing CEI components for the 12 teams shown in the CV chart

# Get unique team data (one row per team/score_type combination)
cv_table_data <- outliers %>%
  # Keep only specific CV bins (matching the chart)
  filter(CV %in% c(0.1, 0.2, 0.25, 0.3, 0.4, 0.5)) %>%
  select(CV, CEI, group_members, score_type) %>%
  distinct() %>%
  arrange(CV, desc(score_type))

# Calculate C and E components for each team
cv_table_data <- cv_table_data %>%
  rowwise() %>%
  mutate(
    ages = list(line_to_vector(group_members, return = "vector") %>% as.numeric()),
    C = compute_CEI(ages, range = c(20, 70), return = "C"),
    E = compute_CEI(ages, range = c(20, 70), return = "E"),
    .keep = "unused"
  ) %>%
  ungroup() %>%
  mutate(
    CV_Group = as.character(CV),
    Type = ifelse(score_type == "max", "High CEI", "Low CEI"),
    CEI = round(CEI, 3),
    C = round(C, 3),
    E = round(E, 3)
  ) %>%
  select(CV_Group, Type, CEI, C, E) %>%
  group_by(CV_Group) %>%
  arrange(desc(Type), .by_group = TRUE)

# Create gt table
cv_table_data %>%
  gt(groupname_col = "CV_Group") %>%
  tab_header(
    title = "Coverage & Evenness Index (CEI) Components",
    subtitle = "Broken down by Coverage (C) and Evenness (E) for the 12 teams shown in CV divergence chart"
  ) %>%
  cols_label(
    Type = "Type",
    CEI = "CEI",
    C = "Coverage (C)",
    E = "Evenness (E)"
  ) %>%
  fmt_number(
    columns = c(CEI, C, E),
    decimals = 3
  ) %>%
  tab_options(
    row_group.as_column = FALSE,
    container.overflow.x = TRUE
  )
```

### Combined Divergence Visualization

Figure 1 below shows, for common values of Blau's Index (Panel A) and the
Coefficient of Variation (Panel B), teams with the highest and lowest CEI.
This highlights that:

- For Blau's Index: CEI can vary substantially (up to 4-fold) even when
  Blau is constant, reflecting different distributions with the same
  binned categories.
- For CV: Particularly for intermediate CV values, CEI can vary widely.
  Towards high CV values, the relationship can reverse as evenness drops.

```{r combined-plot, fig.width=12, fig.height=8}
library(patchwork)
(p_div_blau + labs(title = "")) + (p_div_cv + labs(title = "")) +
  plot_layout(ncol = 2) + plot_annotation(tag_levels = 'A')
```

### Decomposing CEI for Divergent Teams

Table 1 breaks down CEI into its Coverage (C) and Evenness (E) components
for the teams shown in Figure 1. This reveals whether divergences are
driven by differences in spread across the attribute range (Coverage) or
uniformity of distribution (Evenness).

```{r combined-cei-tables}
# Combine both tables with clear section headers
combined_table_data <- bind_rows(
  blau_table_data %>%
    rename(Group = Blau_Group) %>%
    mutate(
      Metric = "Blau Index",
      Group = as.character(Group),
      .before = 1
    ),
  cv_table_data %>%
    rename(Group = CV_Group) %>%
    mutate(Metric = "Coefficient of Variation", .before = 1)
)

combined_table_data %>%
  gt(groupname_col = "Metric", rowname_col = "Group") %>%
  tab_header(
    title = "Coverage & Evenness Index (CEI) Components",
    subtitle = paste(
      "Broken down by Coverage (C) and Evenness (E) for teams",
      "shown in Figure 1"
    )
  ) %>%
  cols_label(
    Type = "Type",
    CEI = "CEI",
    C = "Coverage (C)",
    E = "Evenness (E)"
  ) %>%
  fmt_number(
    columns = c(CEI, C, E),
    decimals = 3
  ) %>%
  tab_options(
    row_group.as_column = FALSE,
    container.overflow.x = TRUE
  )
```

## Sensitivity to range specification

To check how sensitive CEI is to the attribute range supplied, we shift both
the lower and upper boundaries up and down by 50%, 20%, and 0% of the correct
50-year span (20–70). The table shows correlations between CEI values
calculated with each shifted range: Pearson on the lower triangle and
Spearman on the upper triangle.

```{r cei-range-sensitivity, message=FALSE}
range_offsets <- c(-0.5, -0.2, 0, 0.2, 0.5)
range_labels <- scales::percent(range_offsets, accuracy = 1)

# Define shifted ranges while keeping the span constant at 50 years
range_defs <- tibble(
  label = range_labels,
  lower = 20 + range_offsets * 50,
  upper = 70 + range_offsets * 50
)

# Precompute CEI for each range shift
team_ceis <- teams %>%
  group_by(team_id) %>%
  summarise(ages = list(age), .groups = "drop")

cei_variants <- team_ceis
for (i in seq_len(nrow(range_defs))) {
  current <- range_defs[i, ]
  cei_variants <- cei_variants %>%
    mutate(
      !!current$label := purrr::map_dbl(
        ages,
        compute_CEI,
        range = c(current$lower, current$upper)
      )
    )
}

cei_cols <- cei_variants %>%
  select(all_of(range_defs$label))

pearson_corr <- cor(cei_cols, method = "pearson", use = "complete.obs")
spearman_corr <- cor(cei_cols, method = "spearman", use = "complete.obs")

combined_corr <- pearson_corr
combined_corr[upper.tri(combined_corr)] <- spearman_corr[upper.tri(spearman_corr)]

combined_corr %>%
  round(3) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Range shift") %>%
  gt(rowname_col = "Range shift") %>%
  tab_header(
    title = "CEI sensitivity to range specification",
    subtitle = paste(
      "Pearson correlations (lower triangle) and Spearman correlations",
      "(upper triangle) across shifted bounds"
    )
  ) %>%
  tab_options(row_group.as_column = FALSE)
```

## Summary

In this analysis, we generated teams of two different sizes (5 and 10) with
a comprehensive range of member distributions. Five diversity metrics—CEI,
Blau's Index (with 5-year and 10-year binning), Coefficient of Variation
(CV), and Standard Deviation (SD)—were computed for each team.

**Key Findings:**

- **Distribution of Diversity Measures:** Violin plots revealed the
  distribution and spread of each diversity measure across different team
  sizes. While Blau's index is very clustered around a low number of unique
  values, all other distributions are smooth and roughly normal.

- **Correlation Analysis:** The correlation matrix indicated strong
  correlations between certain measures (e.g., CV and SD) while others
  showed weaker relationships. This suggests that while some diversity
  metrics capture similar aspects of diversity, others may provide unique
  insights.

- **Measures Divergence:** A comparison of 7 specific teams highlighted
  that each index can result in a unique ordering, so that the choice
  matters. Similarly, looking at the residuals between CEI and Blau's Index
  or CV, we identified the teams where the measures diverged the most - and
  would argue that in all these cases, CEI indicates an appropriate
  difference between the teams.
